---
title: Is the WASH sector beating the economic curve?
subtitle: Capstone project for b3nharris for ds4owd-001
author: HARRIS Benjamin
format: html
execute: 
  echo: true
  warning: false
editor: visual
editor_options: 
  chunk_output_type: console
---

# project-b3nharris

Capstone project for b3nharris for ds4owd-001

## Is the WASH sector beating the economic curve?

**Concept for the project taken from the blog:** [`https://www.ircwash.org/blog/%E2%80%9Ci-am-bored-work%E2%80%9D-washs-mid-life-crisis`](https://www.ircwash.org/blog/%E2%80%9Ci-am-bored-work%E2%80%9D-washs-mid-life-crisis)

-   Global comparison on progress on poverty (WB data) and WASH (JMP households)
    -   Water and Sanitation seperately
-   Country-level analysis to identify positive outliers (countries making faster progress on WASH than poverty)
-   Analysis on aid flows (OECD) to see if those countries making the best progress are also those receiving most aid per capita (overall and to the WASH sector)
-   Franceys work on GDP: <https://www.franceys.info/blank-2>
-   WaterAid Pub on SE Asia and GDP: <https://washmatters.wateraid.org/publications/achieving-total-sanitation-and-hygiene-coverage-within-a-generation-lessons-from-east>
-   WaterAid: brazil

## Data Sources

```{r}

```

### Poverty

-   **Indicator:** Poverty headcount ratio at \$2.15 a day (2017 PPP) (% of population)

-   **Source:** World Bank, Poverty and Inequality Platform. Data are based on primary household survey data obtained from government statistical agencies and World Bank country departments. Data for high-income economies are mostly from the Luxembourg Income Study database. For more information and methodology, please see http://pip.worldbank.org.

-   **Link:** <https://data.worldbank.org/indicator/SI.POV.DDAY>

-   **Accessed via:** WBData API

### Access to Water

-   **Indicator:** People using at least basic drinking water services (% of population)

-   **Source:** WHO/UNICEF Joint Monitoring Programme ( JMP ) for Water Supply, Sanitation and Hygiene ( [washdata.org](http://washdata.org/) ).

-   **Link:** <https://data.worldbank.org/indicator/SH.H2O.BASW.ZS>

-   **Accessed via:** WBData API

### Access to Sanitation

TBC

## Import Data

```{r}
library(tidyverse)
library(gt)
library(wbstats)
library(imputeTS)
library(janitor)
  
source("WB_data.R")
```

## Global trends on WASH and poverty

Summarise by year to get global averages

```{r}

```
